### 2015-09-16 tex/conclusion.tex
Updated exended paper to reflect VMCAI updates.

Although our approach is inspired by the work in the static analysis
domain, the heap representations used by those techniques do not lend
themselves to symbolic execution since they rely on
approximations, they are inflexible in what types of relations can be
expressed, and they rely on rewriting of constraints.  In this work,
we present a heap summary technique for symbolic execution that
maintains exactness, represents arbitrary heap structures without
restrictions on aliasing relationships, and supports dynamic
on-the-fly initialization of heap structures. This unique
combination of capabilities enables the summary heap to outperform
state-of-the-art symbolic execution techniques, and in some cases, the
performance gains are considerable. We believe that performance gains
in symbolic execution can naturally benefit a variety of analyses
based on symbolic execution, for example, verification of properties,
especially those related to the heap, test case generation, program
evolution techniques such as directed incremental symbolic
execution~\cite{person:pldi2011}, and other back-end analyses.

>>>>

In this work we present a \symtxt{} algorithm for uninitialized
references in a fully symbolic heap. Where GSE lazily instantiates
uninitialized references to either NULL, a new instance of the correct
type, or an alias to a previously initialized object, the symbolic
initialization in this work creates a guarded value set expressing all
of these eventualities in the same symbolic heap. We prove that the
\symtxt{} algorithm is sound and complete with respect to properties
provable by GSE.

An initial implementation of the technique has been done within the
Java PathFinder framework. The \symtxt{} algorithm outperforms
state-of-the-art case-splitting based symbolic execution techniques,
and in some cases, the performance gains are considerable in a set of
Java data structure artifacts. The performance gains in symbolic
execution may naturally benefit a variety of analyses based on
symbolic execution, for example, verification of properties,
especially those related to the heap, test case generation, program
evolution techniques such as directed incremental symbolic
execution~\cite{person:pldi2011}, and other back-end
analyses. Potential future work includes evaluating the impact of
analyses that rely on symbolic execution.

### 2015-09-16 tex/evaluation.tex
Added additional BinarySearchTree test results to table 1.
Updated exended paper to reflect VMCAI updates.

%\begin{figure*}[t]
\begin{table*} [t]
  \centering
  \scalebox{0.88}{\begin{tabular}{| c | c | r | r | r | r | r | r | r | r | r |}
  \hline
   \multirow{2}{*}{Method }&\multirow{2}{*}{ $k$ }
   &\multicolumn{3}{|c|}{Time (seconds)} &\multicolumn{3}{|c|}{States} &\multicolumn{3}{|c|}{ Paths }\\
								&	&\gsetxt{} & L\#		&SL		&\gsetxt{}	& L\# & SL&\gsetxt{}	& L\# 	& SL\\
   \hline
    \multirow{3}{*}{LinkedList }			&3	& 0.91	& 1.21	& 0.69	& 2465	& 2844	& 99		&1656	& 1269	& 25\\
   		 						& 4	& 2.92	& 3.35	& 0.91	& 25774	& 29977	& 155	&17485	& 13550	& 39\\
   								& 5	& 20.78	& 19.47	& 1.59	& 341164	& 400296	& 223	&232743	& 181849	& 56\\
								& 6	& 280.56	& 299.19	& 2.36	&5447980	&6437201	& 303	&3731094	&2933027	& 76\\
								& 7	& -		& -		& 5.07	& -		&-		& 395	&-		&-		& 99\\
								& 8	& -		& -		& 17.49	& -		&-		& 499	&-		&-		& 125\\
								& 9	& -		& -		& 63.96	& -		&-		& 615	&-		&-		& 154\\
								& 10	& -		& -		& 206.93	& -		&-		& 743	&-		&-		& 186\\
    \hline
    \multirow{3}{*}{BinarySearchTree }	& 1	& 0.26	& 0.28	& 0.36	& 19		& 23		& 29		& 6		& 6		& 6\\
   		 						& 2	& 0.83	& 1.28	& 0.93	& 143	& 143	& 145	& 43		& 42		& 33\\
   								& 3	& 20.63	& 25.55	& 4.03	& 1953	& 1703	& 1485	& 515	& 515	& 328\\
    \hline
      \multirow{3}{*}{TreeMap}			& 1	& 0.47	& 0.52	& 0.77	& 65		& 70		& 215	& 11		& 11		& 11\\
   		 						& 2	& 8.99	& 9.73	& 4.72	& 1009	& 942	& 3219	& 127	& 122	& 73\\
   								& 3	& -		& -		& 145.56	& -		& -		& 78695	& -		& -		& 887\\

    \hline
  \end{tabular}}
  \caption{Comparing the efficiency of the summary heap approach to GSE and Lazier\# algorithms.}
  \label{tab:results}
\end{table*}
%\end{figure*}

%\subsection{Analysis}

The results of the experiments are presented in Table
\ref{tab:results}. Each row contains the results for the specified $k$-bound 
for each artifact evaluated. The columns show the total
wall clock time in seconds, states explored, and paths generated for
each of the algorithms. The headings \gsetxt{}, L\#, and SL correspond
to the Generalized Symbolic Execution, Lazier\#, and \symtxt{}
algorithms, respectively. Note that we do no present the results for
the Lazier algorithm since they are almost identical to that of Lazier\#
for our examples. Table entries containing '-' correspond to test runs
that exceeded the allotted time bound of 10 minutes.

In most tests, run times strongly favor the \symtxt{} algorithm, especially for
large $k$-values. Performance improvement ranges from $4.8x$ for
BinarySearchTree at $k$=3, to $118x$ for LinkedList at $k$=6. In fact,
for certain $k$-bounds a number of experiments completed exploration
using the \symtxt{} algorithm in a few seconds whereas \gsetxt{} or Lazier\# were
unable to finish, e.g., BinarySearchTree for $k$=4 and LinkedList for
$k$=8. These results demonstrate the potential of using the \symtxt{} algorithm 
to handle data input for complex types. The lower running times
show that we can prove properties using the \symtxt{} algorithm faster than
\gsetxt{}.

% Some of this discrepancy may be accounted for by the fact that
%\symtxt{} is equipped for incremental solving, however, during
%informal testing with the incremental solver turned off, \symtxt{}
%usually performed better than \gsetxt{} and Lazier\#, especially in
%test runs with high k-bounds.

Path counts likewise substantially favor the \symtxt{} algorithm. The number of
paths explored by the algorithm is strictly less than or equal to the number
of paths explored by \gsetxt{} for all test cases. We can use this
result to do more efficient test case generation. From a path
generated we can use the solutions provided by the constraint solver
to instantiate a set of concrete heaps to provide as test input. These
can provide a smaller test suite to achieve control-flow path coverage when
compared to GSE and Lazier\#.

%Interestingly, \symtxt{} path growth appears
%to be sub-exponential in $k$ for the LinkedList program.

The number of states varies between algorithms, for example, \gsetxt{}
has additional points of nondeterminism during field reads, but in
contrast reference compares are completely deterministic. Thus, in
example programs with large numbers of reference compares, such as
TreeMap, state counts for Lazier\# and the \symtxt{} algorithm may
exceed those for \gsetxt{}. We observed that the additional states
generated by the summary heap algorithm are unsatisfiable at the point
of reference compares; this is why they do not contribute any
additional branches in the final symbolic execution tree. Furthermore,
the larger state count does not impact the overall runtime of the
technique.

%The additional states generated by summary heap are unsatisfiable
%which is why they do not add nodes in the final symbolic execution
%tree. Furthermore, the larger state count does not impact the overall
%runtime of the technique.

% Despite this, in practice this is often mitigated by the
%fact that \symtxt{} can dispatch address compares without consulting
%the constraint solver.

In summary, the benefits of avoiding
initialization-based non-determinism outweigh the increased
complexity in the constraints due to the advances made in SMT
solvers in these examples. In fact, the \symtxt{} algorithm can analyze certain types of programs with
orders of magnitude greater efficiency than with \gsetxt{} or
Lazier\#, while covering exactly the same control flow paths.

>>>>

\section{Evaluation}
\label{sec:evaluation}
%\begin{figure*}[t]
\begin{table*} [t]
  \centering
  \caption{Comparing \symtxt{} with the~\gsetxt{} and Lazier\# algorithms.}
  \scalebox{0.88}{\begin{tabular}{| c | c | r | r | r | r | r | r | r | r | r |}
  \hline
   \multirow{2}{*}{Method }&\multirow{2}{*}{ $k$ }
   &\multicolumn{3}{c|}{Time (seconds)} &\multicolumn{3}{c|}{States} &\multicolumn{3}{c|}{ Paths }\\
								&	&\gsetxt{} & L\#		&SL		&\gsetxt{}	& L\# & SL&\gsetxt{}	& L\# 	& SL\\
   \hline
    \multirow{3}{*}{LinkedList }			&3	& 0.91	& 1.21	& 0.69	& 2465	& 2844	& 99		&1656	& 1269	& 25\\
   		 						& 4	& 2.92	& 3.35	& 0.91	& 25774	& 29977	& 155	&17485	& 13550	& 39\\
   								& 5	& 20.78	& 19.47	& 1.59	& 341164	& 400296	& 223	&232743	& 181849	& 56\\
								& 6	& 280.56	& 299.19	& 2.36	&5447980	&6437201	& 303	&3731094	&2933027	& 76\\
								& 7	& -		& -		& 5.07	& -		&-		& 395	&-		&-		& 99\\
								& 8	& -		& -		& 17.49	& -		&-		& 499	&-		&-		& 125\\
								& 9	& -		& -		& 63.96	& -		&-		& 615	&-		&-		& 154\\
								& 10	& -		& -		& 206.93	& -		&-		& 743	&-		&-		& 186\\
    \hline
    \multirow{3}{*}{BinarySearchTree }	& 1	& 0.26	& 0.28	& 0.36	& 19		& 23		& 29		& 6		& 6		& 6\\
   		 						& 2	& 0.83	& 1.28	& 0.93	& 143	& 143	& 145	& 43		& 42		& 33\\
   								& 3	& 20.63	& 25.55	& 4.03	& 1953	& 1703	& 1485	& 515	& 515	& 328\\
								& 4	& -		& -		& 410.89	& -		&-		& 73635	&-		&-		& 15563\\
    \hline
      \multirow{3}{*}{TreeMap}			& 1	& 0.47	& 0.52	& 0.77	& 65		& 70		& 215	& 11		& 11		& 11\\
   		 						& 2	& 8.99	& 9.73	& 4.72	& 1009	& 942	& 3219	& 127	& 122	& 73\\
   								& 3	& -		& -		& 145.56	& -		& -		& 78695	& -		& -		& 887\\

    \hline
  \end{tabular}}

  \label{tab:results}
\end{table*}

%\end{figure*}

%\subsection{Analysis}
The \symtxt{} algorithm is implemented as an extension to the Symbolic
PathFinder (SPF)~\cite{DBLP:journals/ase/PasareanuVBGMR13}
framework. In addition to the operations presented in this paper, the
implementation contains support for operations over integers,
calculating per-path preconditions and postconditions, as well as
generating test input heaps that exercise all feasible control flow
paths. Future work is adding support for floating point
operations, arrays, and bit-operations.

The \symtxt{} implementation uses \texttt{jConstraints} with the z3 solver \cite{ase2014-ghilrr,jpf2014-dghirr,deMouraBjorner08Z3}. The implementation takes advantage of incremental solving and employs caching for performance since only small portions of the heap constraint changes during the search.

SPF includes an implementation of~\gsetxt{} with lazy initialization.
In recent work, we implemented the Lazier and Lazier\# algorithms in
SPF~\cite{Hillery:2014}; these constitute the state of art approaches
case-splitting based lazy initialization techniques. The goal of our
experiment is to evaluate the efficacy of our approach with respect to
other state-of-the-art techniques for symbolic execution of programs
with unbounded complex data input. We empirically evaluate the
\symtxt{} algorithm by considering the following research question:
How does the cost of the \symtxt{} algorithm (SL) compare with that of
\gsetxt{} (GSE), Lazier, Lazier\# (L\#) algorithms?

The \emph{independent} variable in our study is the $k$-bound;
$k$-bounding bounds the length of a reference chain from the root of
the heap~\cite{Kiasan06}.  We select three dependent variables and
measures for our study: (i) time, (ii) states explored, and (iii)
paths generated. The \emph{time} is the total time taken by each
algorithm to explore the symbolic execution tree as well as total
constraint solving time. The \emph{states explored} represents the
number of nodes in the symbolic execution tree, and the \emph{paths
  generated} represents the number of unique paths in the symbolic
execution tree.

The data structures evaluated are a standard set commonly used in
analyses involving heap-manipulating
programs~\cite{Deng:2006,Deng:2007,boyapati2002korat,Ferrara:2014,Rosner:2015},
including a linked list, binary search tree, and red/black tree. We
use a repOk() method (a class invariant) for data structures in
object-oriented code~\cite{boyapati2002korat}. The repOk() methods
provide a mechanism to generate valid inputs for the methods under
test; symbolic execution has been used to generate test inputs using
repOk() methods in~\cite{visser2004test}. In this work, we define and
use repOk() methods to generate valid heaps with the appropriate
$k$-bounds before executing the methods under test. Note that this
allows us to have precision in checking properties of the heap that is
not possible in static analysis based techniques~\cite{Dillig:2011}.

The results of the experiments are presented in
Table~\ref{tab:results}.  Each row reports the results for the
specified $k$-bound for each artifact evaluated. The columns show the
total time in seconds, states explored, and paths generated for each
algorithm. The headings \gsetxt{}, L\#, and SL correspond to the GSE,
Lazier\#, and \symtxt{} algorithms, respectively. The results for the
Lazier algorithm are omitted because they are nearly identical to the
Lazier\# results for these artifacts. A table entry of '-' indicates
the analysis exceeded the allotted time bound of 1 hour.

The number of possible non-isomorphic heap configurations grows
exponentially for case-splitting techniques with a monotonic increase
in $k$, resulting a corresponding exponential increase in analysis
times . This is evident in all the examples in
Table~\ref{tab:results}. The GSE and L\# algorithms are unable to
finish exploration in a time bound of one hour for $k\geq 7$ for the
LinkedList, $k\geq 4$ for the BinarySearchTree, and $k\geq 3$ for the
TreeMap examples. The improvement in analysis time for the \symtxt{}
over the state of the art case splitting techniques range from $4.8x$
for BinarySearchTree at $k$=3, to $118x$ for LinkedList at $k$=6. In
fact, for some $k$-bounds a number of experiments complete exploration
using the \symtxt{} algorithm in a few seconds whereas \gsetxt{} or
Lazier\# are unable to finish, e.g., BinarySearchTree for $k$=4 and
LinkedList for $k$=8.

The number of path explored by \symtxt{} algorithm are strictly less
than or equal to the number of paths explored by \gsetxt{} for all the
evaluated artifacts. We can use this result to do more efficient test
case generation. From a generated path and corresponding symbolic heap
we can use the solutions provided by the constraint solver to
instantiate a set of concrete heaps to provide as test input. These
provide a smaller test suite to achieve control-flow path coverage as
compared to GSE and Lazier\#. Note that the ability to perform test
input generation is another advantage of \symtxt{} technique over
static analysis approaches.

The number of states varies between algorithms, for example, \gsetxt{}
has additional points of nondeterminism during field reads, but in
contrast reference compares are completely deterministic. Thus, in
example programs with large numbers of reference compares, such as
TreeMap, state counts for Lazier\# and the \symtxt{} algorithm may
exceed those for \gsetxt{}. We observed that the additional states
generated by the summary heap algorithm are unsatisfiable at the point
of reference compares; this is why they do not contribute any
additional branches in the final symbolic execution tree. Furthermore,
the larger state count and the corresponding satisfiability checks on
the constraint solver does not increase the overall runtime of the
technique.

In summary, the benefits of avoiding case-splitting based
non-determinism outweigh the increased complexity in the constraints
over heaps due to the advances made in SMT solvers in these
examples. In fact, the \symtxt{} algorithm can analyze certain types
of programs with orders of magnitude greater efficiency than with
\gsetxt{} or Lazier\#, while covering exactly the same feasible
control flow paths in the program.

### 2015-09-16 tex/lazy.tex
Updated exended paper to reflect VMCAI updates.

\section{Generalized Symbolic Execution}

Recall that in generalized symbolic execution (GSE) for lazy
initialization of symbolic references consists of three key points of
non-determinism where each symbolic reference can be initialized
non-deterministically to null, a new instance of the symbolic
reference, or aliases to symbolic references of the same type
previously initialized. The initialization in GSE consists of creating
branches in the execution tree for all the non-deterministic choices.

>>>>

\section{GSE}
\label{app:gse}
This section introduces \gsetxt\ semantics on which \symtxt\ builds. GSE and its variants have the same non-deterministic choice given a reference to an uninitialized object; that reference can point to null, a new instance of an object with the correct type, or to an object of the same type that has been instantiated previously by GSE \cite{GSE03,KiasanKunit,Cadar:2008,Rosner:2015}. The objects instantiated for new instances are referred to as the \emph{input heap}. Only these objects comprise the potential alias set when GSE encounters references to uninitialized objects. In general, the GSE search space branches at uninitialized objects, and the number of branches depends on the size of the input heap.

The lazy initialization rules for GSE on the symbolic heap are in
\figref{fig:lazyInit}: NULL, new, and alias. The rules rely on fresh references and locations that strictly increase so it is possible to minimize over a set to find the first created. Further, references are partitioned to support latter proofs: $\mathrm{init}_\cfgnt{r}()$ for the input
  heap
  ; $\mathrm{fresh}_\cfgnt{r}()$ for \emph{auxiliary
  literals}
  ; and $\mathrm{stack}_\cfgnt{r}()$ for \emph{stack
    literals}.
  In general, as shown in the next section,
  only input heap references appear in constraints to express potential aliasing, and only stack references appear in environments, expressions, or continuations.  Finally, the \emph{isInit} function is true for initialized references from the input heap (i.e., potential aliases).

How the lazy
initialization is used is defined in \figref{fig:lazy} with the $\rgse$ relation collecting all the rules into a single relation on states. This relation is used later by the proof. GSE initialization takes place on the field-access rule in \figref{fig:lazy},
using the $\rightarrow_I^*$ relation from \figref{fig:lazyInit}, to ensure the accessed field is
instantiated. Initialization in GSE never
happens for more than one object on any use of $\rightarrow_I^*$:
the set $\Lambda$ is either empty or contains exactly one location. This property is an artifact of how GSE case splits when it instantiates: each choice is a new unique heap. This changes in the next section with the new symbolic initialization that collects all the choices into a single heap using guarded value sets.

The rest of the rules in \figref{fig:lazy} do not initialize, but they are included to elucidate how symbolic initialization differs from GSE with lazy initialization. In particular, there is no branching in the search space on reference compare for GSE because references point to a single location after initialization. The new symbolic initialization using guarded value sets in this papers changes this behavior.

### 2015-09-16 tex/rw.tex
Updated exended paper to reflect VMCAI updates.

%\subsection{Other Heap Representations in Symbolic Execution}
%Need to talk about GSE, DSE, SAGE, PEX
%GSE
Symbolic initialization also draws inspiration from lazy initialization, the core idea of generalized symbolic execution (\gsetxt{})\cite{GSE03}. Several projects have used lazy initialization to conduct symbolic execution on programs with more general types,
including references and
arrays~\cite{KiasanKunit,Cadar:2008,Deng:2007,Rosner:2015,Blackshear:2013,Filieri:2015}. However, \gsetxt{} techniques branch over
multiple copies of the system state during dereferencing operations,
exacerbating path explosion.

@@

Several separation logic solvers have been proven to be sound and complete for heaps with linked lists~\cite{Navarro:2011,Cook:2011,Berdine:2005}, trees~\cite{Piskac:2014}, or data structures satisfying user-supplied invariants~\cite{Brotherston:2014}. Separation logic solvers can also reason about entailment which is a subject of future work for symbolic initialization. Solutions in separation logic do not lend themselves to test case generation is  goal in this work.

>>>>

%\subsection{Other Heap Representations in Symbolic Execution}
%Need to talk about GSE, DSE, SAGE, PEX
%GSE
Symbolic initialization also draws inspiration from lazy initialization, the core idea of generalized symbolic execution (\gsetxt{})\cite{GSE03}. Several projects have used lazy initialization to conduct symbolic execution on programs with more general types,
including references and
arrays~\cite{KiasanKunit,Cadar:2008,Blackshear:2013,Filieri:2015}. Improvements to the basic lazy initialization algorithm have been proposed, including delaying aliasing choices~\cite{Deng:2007}, or checking initializations against invariants as they occur~\cite{Braione:2015,Rosner:2015}. However, these \gsetxt{} techniques branch over
multiple copies of the system state during dereferencing operations,
exacerbating path explosion.

@@

Several separation logic solvers have been proven to be sound and complete for heaps with linked lists~\cite{Navarro:2011,Cook:2011,Berdine:2005}, trees~\cite{Piskac:2014}, or data structures satisfying user-supplied invariants~\cite{Brotherston:2014}. Separation logic solvers can also reason about entailment which is a subject of future work for symbolic initialization. Solutions in separation logic do not lend themselves to test case generation,  a goal of this work.

### 2015-09-16 tex/summary-heap.tex
Updated exended paper to reflect VMCAI updates.

\section{A Symbolic Heap}
A \emph{symbolic heap}, as defined in this work, is a tuple
$(\cfgnt{L}\ \cfgnt{R}\ \phi\ \eta)$, indicating a \emph{location map}
$\cfgnt{L}$, a \emph{reference map} $\cfgnt{R}$, a \emph{path condition} $\phi$, and an \emph{environment} $\eta$.  Given a reference $r$ in the heap,
$\cfgnt{L}(r) = \{(\phi\ \cfgnt{l})\ ...\}$ is the guarded value set
associated with $r$. The constraint associated with each
location is a guard representing the conditions under which the
associated reference maps to the given location. By definition, any
location appears at most once in any given value set. Unions of value
sets containing the same location are resolved by forming a
disjunction of the constraints from each set. $\cfgnt{R}(\cfgnt{l},
\cfgnt{f}) = \cfgnt{r}$ is the reference associated with the given
location-field pair in the heap.

>>>>

\section{A Symbolic Heap}
\label{sec:symheap}
A \emph{symbolic heap}, as defined in this work, is a tuple
$(\cfgnt{L}\ \cfgnt{R}\ \phi\ \eta)$, indicating a \emph{location map}
$\cfgnt{L}$, a \emph{reference map} $\cfgnt{R}$, a \emph{path condition} $\phi$, and an \emph{environment} $\eta$.  Given a reference $r$ in the heap,
$\cfgnt{L}(r) = \{(\phi\ \cfgnt{l})\ ...\}$ is the guarded value set
associated with $r$. The constraint associated with each
location is a guard representing the conditions under which the
associated reference maps to the given location. By definition, any
location appears at most once in any given value set. Unions of value
sets containing the same location are resolved by forming a
disjunction of the constraints from each set. $\cfgnt{R}(\cfgnt{l},
\cfgnt{f}) = \cfgnt{r}$ is the reference associated with the given
location-field pair in the heap.
